---
title: "Solving ODEs in the wild: Scalable pharmacometrics with Stan"
author: "Sebastian Weber, <sebastian.weber@novartis.com>"
date: "20th May 2018"
output: html_vignette
bibliography: stancon18.bib
---

```{r cache=FALSE, echo=FALSE}
library(knitr)
knitr::read_chunk("warfarin_pk.R")
knitr::read_chunk("warfarin_pd.R")
knitr::read_chunk("warfarin_pd_tlagMax_1.stan", labels="stan-pd-model-1", from=6, to=20)
knitr::read_chunk("warfarin_pd_tlagMax_2.stan", labels="stan-pd-model-2", from=6, to=20)
knitr::read_chunk("warfarin_pd_tlagMax_2par.stan", labels=c("stan-pd-parmodel-fun"), from=c(20), to=40)
knitr::opts_chunk$set(echo=FALSE,fig.width=1.62*5,fig.height=5,cache=TRUE)
```

# Abstract

Pharmacometric modeling involve nonlinear hierarchical models which
are most naturally expressed as ordinary differential
equations. These class of models lead to a number of challenges
which complicate a practical modeling work-flow in Stan. At the
example of the public Warfarin data-set I will present solutions to
these. I will first illustrate the concept of forcing functions and
how to use these efficiently in Stan. Moreover, I will show how the
hierarchical structure of the problem is most efficiently taken
advantage of to parallelize the model computations which expedites
the model evaluation time and allows to scale Stan's performance to
large data-sets given sufficient computing resources.

# Introduction

A drug therapy aims to treat a disease state. The drug is administered
through some route in order to reach some location in the human body
where the drug is active. The the human body is exposed to the drug
and a readily accessible measure for drug exposure is the drug
concentration in blood which is often taken as surrogate for exposure
at the target tissue. The key to a successful therapy is to attain
quickly an adequate exposure which is high enough for a sufficient
effect to be reached, but low enough to avoid undesired adverse
affects. This defines the so-called therapeutic window within which
one aims to maintain patients using an adequate dosing regimen.

Pharmacometrics now aims to model the longitudinal history of a
patient being under drug treatment. That is, one models how drug
intake translates into a drug concentration (pharmacokinetics) and how
drug concentration drives some effect (pharmacodynamics) over
time. These processes are most naturally stated as ordinary
differential equations (ODEs) where the phamacokinetics (PK) follows
mass-action kinetics and the pharmacodynamics (PD) is modeled using a
link process relating drug concentration (PK) with drug effect (PD).

This translates mathematically into a forced ODE problem. The PK model
describes the so-called absorption, distribution, metabolism and
elimination process (ADME). The PK model itself is externally driven
by the input into the system which is the intake of drug amount over
time. The intake of drug amount is referred to as dosing regimen and
is known as data in advance or recorded during the trial. The PD model
then describes the effect of the drug concentration on some measure of
efficacy or safety.

Oftentimes these problems are approached in a two-step
manner. First the PK model is informed using blood concentration
data. Then the model parameters are fixed and the PD model is
informed. Thus, the PK model becomes a forcing function to the PD
model. The "clean" Bayesian solution is a joint fit of the
combined model, but oftentimes a two-step approach is much quicker
to implement and it gives a great example for forcing functions
here.

In the following I will first introduce the warfarin data-set, then
illustrate the concept of forcing functions and how to introduce
them efficiently in Stan. Finally, I will show how the hierarchical
problem structure can be used most efficiently to parallelize
computations and expedite the model evaluation time for the example
at hand, but most importantly enable Stan to scale it's performance
to large data sets when given sufficient computing resources.

## Anticoagulant Warfarin

### Pharmacokinetics

Warfarin is an anticoagulant (blood thinner) used to prevent strokes
in certain conditions. The pharmacokinetics of Warfarin
[@OReilly1963,@Holford1986] is well-described by a one compartmental
model which states that drug amounts (mass) is cleared from a
compartment via a first order process,

$$ d(C_i(t) \, V_i)/dt = - Cl_i \, C_i(t) + f_{in,i}(t).$$ 

Here the index $i$ labels patients and the quantities introduced are

- $C_i(t)$: Drug concentration in the central compartment at
time-point $t$.
- $V_i$: Apparent volume of the central compartment.
- $Cl_i$: Clearance of the drug from the central compartment in units
of mass per volume.
- $f_{in,i}(t)$: Uptake of drug into the system which is determined by
  the route of administration.

Note that while we measure *concentrations* $C_i(t)$ of some patient
$i$, the mass-action kinetics holds for *mass*. Moreover, the uptake
of drug may occur through various routes like intra-venous injections,
intra-venous infusions or oral administration. The example data set
studied a single orally administered Warfarin dose of 1.5mg/kg. Oral
administration is usually well-described by a first order process
itself (input rate is proportional to $k_{a,i}$) such that the full PK
system is most conveniently written as

$$ d(a_i(t))/dt = - k_{a,i} \, a_i(t)$$
$$ d(C_i(t) \, V_i)/dt = - Cl_i \, C_i(t) + k_{a,i} \, a_i(t).$$ 

Fortunately, this ODE system can solved analytically. After
administration of a single dose at $t=0$ the solution is 

$$ C_i(t) =  \frac{D_i}{V_i} \, \frac{k_{a,i}}{k_{a,i} - Cl_i/V_i}
\left( \exp{(-Cl_i/V_i \, t)} - \exp{(-k_{a,i} \, t)}\right).$$

However, Warfarin has (as many drugs) an additional important property
which is a delay of the drug-uptake. Thus, administering the drug at
$t=t_0$ does not cause an immediate diffusion of the drug to the
central compartment, but rather it takes some amount of time $t_{lag}$
to see a non-zero concentration in the main compartment. The need for
a time lag suggests that the actual absorbtion process is likely more
involved than a simple first process, but a lag time is adequate here
to account for it given that the interest is in time-scales much
larger than the lag time. One approach to account for the lag time is
to infer $t_{lag}$ and then shift the analytic solution without a lag
time by this amount. So the starting time of drug uptake is shifted by
$t_{lag}$. While common in practice, this introduces the need for a
discontinuous if-statement which depends on a parameter. This is not
ideal for the performance of any gradient based approach like Stan's
NUTS.

The data set we consider comprises 32 patients. For 14 patients blood
samples measuring the concentration in the central compartement have
been taken starting from 0.5h post oral dose administration while for
18 patients sampling started at 24h post oral dose administration. The
hierarchical model for the parameters is

$$ \mbox{logit}^{-1}(t_{lag,i}/t_{lag,max}) \sim N(\nu_1, \sigma_1^2)$$
$$ \log(k_{a,i}) \sim N(\nu_2, \sigma_2^2) $$
$$ \log(Cl_i) \sim N(\nu_3 + 3/4 \, \log(wt_i/70kg), \sigma_3^2)$$
$$ \log(V_i) \sim N(\nu_4 + \log(wt_i/70kg), \sigma_4^2).$$

The model takes allometric scaling into account for the clearance and
volume. This scaling accounts for a size dependent metabolic activity
of the organs involved in eliminating the drug and the slopes are
derived from first principles. The fit of the PK model is shown below
for 4 patients as an example and for the full population. All codes
are provided in this notebook, but in the following we will
concentrate on the pharmacodynamic model of Warfarin. The parameters
of the PK model are in the following assumed known and they are set to
the median values of the PK model fit.


```{r pk-init,include=FALSE}
```

```{r pk-post-4}
```

```{r pk-post-pop}
```

### Pharmacodynamics

```{r pd-init,include=FALSE}
```

The effect of Warfarin is to thin blood and a quantitative measure is
its effect on the prothrombin complex levels in relation to normal
levels.

In the following, the pharmacodynamics will be conditioned on the
pharmacokinetic model and since we have fixed the model parameters,
the PK model is effectively a forcing function to the PD model. Thus,
the concentration $C_i(t)$ at time $t$ for patient $i$ can be treated
as data within the context of the PD model. An appropriate PD model for
the effect of Warfarin on prothrombin complex levels is the turn-over
model. The turn-over model is semi-mechanistic since it is certainly
an over-simplification of the biological processes, but it is still
motivated by biological rationale. Specifically, the response $R_i(t)$
is considered to be represented by a compartment which has a
zero-order influx $k_{in,i}$ and a first order out-flux
$k_{out,i}$. The drug-effect may enter this model through an
inhibition or stimulation of either process. For Warfarin
[@OReilly1968,@Holford1986], an inhibition of the zero-order $k_{in}$
is an adequate choice,

$$ dR_i/dt = k_{in,i} \, (1 - \frac{C_i(t)}{C_i(t) + EC50_i}) - k_{out,i} \, R_i $$
$$ \Leftrightarrow dR_i/dt = k_{in,i} \, \left\{1 - \mbox{logit}^{-1}[\log(C_i(t)) - \log(EC50_i)]\right\} - k_{out,i} \, R_i. $$

Whenever the drug concentration is equal to the $EC50_i$ value, then
half of the full drug effect is reached. Furthermore, in absence of
drug, the response $R_i$ will reach a steady-state which is $R_{i,ss}
= \frac{k_{in,i}}{k_{out,i}}$ (just set $dR_i/dt=0$). Since at $t=0$
the human body is free of any drug and we assume that at $t=0$ the
system is in steady-state such that the response at $t=0$ is used to
set $k_{in,i}$ by multiplication with $k_{out,i}$ (so we do not fit
$k_{in}$ as an independent parameter). The hierarchial model structure
for the PD model is

$$ \log(R_{0,i}) \sim N(\theta_1, \omega_1^2)$$
$$ \log(1/k_{out,i}) \sim N(\theta_2, \omega_2^2)$$
$$ \log(EC50_{i}) \sim N(\theta_3, \omega_3^2).$$

In this form, the turn-over model describes a reduction of the
response variable as a result of drug treatment, since the value of
$k_{in}$ is decreased through the drug. This is in line with the raw
data:

```{r pd-raw-data-plot}
```

The ODE functor which defines the turn-over model could look like
shown below when coded in Stan:

```{r stan-pd-model-1,echo=TRUE,eval=FALSE}
```

It is a good practice to visualize the ODE solution with typical
values (for example those values used to define prior means) before
using the ODE functor inside Stan. In R this can be achieved through
the use of `deSolve` and `expose_stan_functions`:


```{r pd-ode-check,echo=TRUE}
```

This shape of the function roughly resembles the raw data. However,
the time to fit this relatively small data set is rather excessive:

```{r pd-stan-fit-1,include=FALSE}
```

```{r pd-stan-fit-1-time}
```

In words, it takes for only 32 patients about 15 minutes for 250
warmup and 250 sampling iterations on a modern machine used inside a
high-performance cluster.

The above formulation of the problem tickles a few issues of the
current Stan language which we can avoid. In the current Stan language
all variable definitions inside a function are automatically
considered as parameters if the output of the function is involving
parameters of the model, which happens whenever any of the arguments
of a function is a parameter. Recall, that the major numerical cost in
any Stan program is the gradient calculation of the log-likelihood wrt
to all parameters. Inspecting the ODE functor above reveals that all
the parameters of the PK model are implicitly turned into parameters
due to their declaration in a function which returns a result
involving parameters. So, let's reformulate the ODE functor without
these intermediate declarations as shown below:

```{r stan-pd-model-2,echo=TRUE,eval=FALSE}
```

Doing so leads to less readable code, but it does avoid that the PK
model is evaluated with variables which are considered as
parameters. In effect, the evaluation of the PK model is much cheaper
within this ODE functor. Running now the same model with this modified
ODE functor yields a more than doubling of the execution speed:

```{r pd-stan-fit-2,include=FALSE}
```

```{r pd-stan-fit-2-time}
```

Fortunately, the Stan 3 language proposal currently worked on will
allow to deliberately declare which variables are data for sure and
which ones not. Thus, the rather cumbersome second version shouldn't
be needed in the future with Stan 3.

Finally, here are the posterior predictive distributions for the first
4 patients in the data set and below the posterior predictive for the
population with the raw data of all patients.

```{r pd-post-4}
```

```{r pd-post-pop}
```

## Scalable Pharmacometrics

The example presented so far comprises a very small data set and used
on of the simplest PD models. While the presentation so far is no
over-simplification, the major issue is the inability to scale Stan's
performance to larger data sets. However, pharmacometric models are by
default hierarchical such that patients are modeled exchangeable to
one another. Thus the likelihood of a given patient can be evaluated
in independence of all other patients. This hierarchical structure can
be taken advantage of using the new `map_rect` facility in Stan. This
function applies a user-defined defined function to a set of
parameters which are in rectangular data storage format (hence the
name of the function). Since each evaluation is independent of all
others, these evaluations can be performed in parallel using either
threading or the message passing interface (MPI). Threading has less
operating system requirements in comparison to MPI, but is limited to
the CPUs of the machine a given chain is run on while MPI can be used
across multiple machines (typically part of a large high-performance
cluster). The `map_rect` facility is used most efficiently if the
function evaluated per subject (or whatever unit) directly returns the
log-probability density contribution of the respective subject. This
minimizes the communication between the different processes.

```{r pd-parallel,include=FALSE}
```

The function for the Warfarin example could look like this:

```{r stan-pd-parmodel-fun,echo=TRUE,eval=FALSE}
```

In the `model` block this can then be called via:

```{r echo=TRUE,eval=FALSE}
   target += map_rect(lpdf_subject, phi, Eta, x_r, x_i);
```

The speedup for this specific example is shown below when using
threading or MPI on a machine with upto 16 cores.

```{r pd-parallel-wallclock}
```

As can be seen, the total wallclock time can be reduced to less than 1
minute when using 16 cores. Moreover, the MPI backend outperforms the
threading backend considerably for a single core run and in terms of
better scalability. Currently, the thread implementation in Stan
relies on C++11 programming language standards. These do defer many
implementation details to the compiler used. This will hopefully leave
some room for future improvements of the code. From these experiments
MPI should be preferred over threading despite the larger burden to
setup MPI in comparison to threading.

What is important to emphasize is that the `map_rect` approach enables
to **scale** the performance of a given Stan program and data
set. This scalability allows to adapt performance needs in such a way
that we get reasonable run times needed for a productive, iterative
explorative modeling process (assuming availability of sufficient
hardware ressources). As a rule of thumb, hierarchical models with a
large computation cost per unit, like ODE models, greatly benefit from
the use of `map_rect`. For a large problem involving $>1300$ patients
speedups of up to 61x on 80 cores haven been measured (2.5 days down
to 1h computation time) [@Weber2018].

# Conclusion

Stan has come a long way and has as of today all required components
for application to realistic pharmacometric problems. These are
numerically very challenging and have influenced the Stan development
in some aspects considerably (non-stiff/stiff ODE solvers, matrix
exponential and now within-chain parallelization). With the recent
addition of within-chain parallelization Stan can finally be used as a
practical tool for iterative modeling with realistic data-sets as used
in pharmacometrics. The `map_rect` function is currently somewhat
clunky to use due to the need for packing and unpacking of parameters
and data, but the additional work is well offset by the massive
performance gain for large problems whenever sufficient CPU cores are
available to the Stan modeler. Once Stan 3 introduces closures for
functions, the in-convenient packing/unpacking coding will hopefully
be remedied to some extent.

# References
