---
title: "Dynamic Soccer models: checks using simulated data"
output:
  pdf_document: default
toc: true
toc_depth: 3
---

# Summary

This R-notebook checks whether the dynamic Skellam model has been properly implemented in Stan.
It does so by simulating data from the data generating process, and checking if the fitted model 
parameter estimates are similar to the true parameter values used to simulate the data.


```{r, results = 'hide', message = FALSE, warning = FALSE}
rm(list = ls())
library(data.table)
library(ggplot2)
library(rstan)

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
source("code/simulateData.R")
source("code/plot_ground_truth_vs_estimate.R")
source("code/ppc_coverage_plot.R")
source("code/MakeTimeSeriesPlot.R")
source("code/Create_model_data_for_TS2.R")
source("code/addTeamIds.R")
source("code/create_league_table.R")
source("code/MakeEPLPlotAbility.R")
source("code/games_predicted_vs_actual_intervals.R")
source("code/ppc_coverage_plot.R")
source("code/calc_rps_scores.R")
source("code/rankProbScore.R")
source("code/odds_to_probability.R")
source("code/ReadfitsandCalculateRPS.R")
source("code/FitOneStepAhead.R")

# create output folder
if(!dir.exists("output")) {dir.create("output")}

```

# Simulate data acc to dynamic Skellam model

```{r}
source("code/simulateData.R")
my_par_list <- list(nseasons = 2,
                 nteams = 18,
                 nrounds = 34,
                 offense_sigma = 0.25,
                 defense_sigma = 0.25,
                 mixing_proportion = 0.05, # 5% excess draws to test the zero inflation
                 mu_const = 0.2,
                 home_advantage = 0.4,
                 hyper_variance = 0.1,
                 art_turf_vec = c(rep(0, 12), rep(1, 6)), # not used
                 art_turf_advantage = 0 # not used
)


model_data <- Simulate_dynamic_ZPD_data(seed = 123, my_par_list)
```

`Simulate_dynamic_ZPD_data()` first simulates for each team two time series, one for its attack ability, and one for its defense ability.
It then uses these abilities to generate match outcomes for random pairings of teams. The match outcomes are simulated according to the Skellam model (difference of two Poisson distributions), with a zero inflation component added.

# plot the true (simulated) team abilities

```{r warning = FALSE, message = FALSE}
ma_offense <- data.table(melt(model_data$a_offense))

ggplot(ma_offense[variable %in% c(paste("V", 1:(34*2), sep = '')),], aes(x= variable, y = value, 
                       group = team_id, col = team_id)) + geom_line() + facet_wrap(~ team_id, ncol = 4)
```

```{r}
ma_defense <- data.table(melt(model_data$a_defense))

ggplot(ma_defense[variable %in% c(paste("V", 1:(34*2), sep = '')),], aes(x= variable, y = value, 
                       group = team_id, col = team_id)) + geom_line() + facet_wrap(~ team_id, ncol = 4)

  
```

# Fit dynamic Skellam model on simulated data

```{r}
fullrun <- 1

if(fullrun) {
stanfit_sim_skellam <- stan(
  file = "models/skellam_dynamic.stan",
  data = model_data,
  chains = 4, 
  warmup = 200, 
  init_r = 0.1, # instead of 2
  iter = 500,
  cores = 4,
  control = list(adapt_delta = 0.95)
  )
  saveRDS(stanfit_sim_skellam, "FITS/skellam_dynamic_sim.rds")
} 
```

# Analyse results

```{r}
stanfit_sim_skellam <- readRDS("FITS/skellam_dynamic_sim.rds")
print(stanfit_sim_skellam, c("constant_mu", "home_advantage", "mixing_proportion"))
```

# Compare the estimated team abilities with the true team abilities

```{r}
sims <- extract(stanfit_sim_skellam)

a_sims <- sims$a_offense

id_lut <- model_data$id_lut

MakeTimeSeriesPlotSim(a_sims, model_data$a_offense, id_lut, title = "Simulated Attack parameter")
```

```{r}
source("code/MakeTimeSeriesPlot.R")
sims <- extract(stanfit_sim_skellam)

a_sims <- sims$a_defense

id_lut <- model_data$id_lut

MakeTimeSeriesPlotSim(a_sims, model_data$a_defense, id_lut, title = "Simulated defense parameter")
```
