---
title: "Aggregate random coefficients logit—a generative approach"
author: "Jim Savage and Shoshana Vasserman"
date: "17 March 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

# Aggregate random coefficients logit: Bayesian estimation using Stan

This note illustrates how to fit aggregate random coefficient logit models in Stan, 
using Bayesian techniques. It's far easier to learn and implement than the standard BLP algorithm, and has the
benefits of being robust to mismeasurement of market shares, and giving limited-sample posterior uncertainty
of all parameters (and demand shocks). This comes at the cost of modeling firms' price-setting process, including how unobserved product-market demand shocks affect prices. 


## Introduction

A common problem in applied economics, especially in industrial organization or marketing, is that
a decision maker wants a good model of how individual customers make purchase decisions, 
but has data only at the aggregate level.  

Let's give examples of the sort of problem we're trying to solve. 

### Example 1: Regulating a merger

First is the classic merger problem: a regulator is be interested in whether a merger will 
hurt customers, as might be the case if

1. The merging firms offer similar products that appear to be in close competition, and few other options exist. And
2. Marginal economies of scale from the merger are small. 

Let's focus on the first problem. The regulator really needs to know whether the customers of the two firms' products perceive the products as being similar---that is, in genuine competition with one another. 
This is a harder task than it might appear on the surface, as many goods might look similar to an 
outside regulator but are really quite different. A mid-range Mercedes Benz might have similar 
specifications to a Toyota, but is perceived by customers as being a different product. 

Ideally the regulator would have sales-level data for each customer, their purchase decisions, and second choices. But this might be impossible to get. Yet it is quite straightforward to purchase aggregate sales 
data at the product-market level from market research firms. So the regulator has to make do with that. 

### Example 2: A manager considering a new product or a new market

Managers ideally want to create goods and products where there is a strong latent demand but
little competition. To do this, they need to understand the distribution of customers' preferences 
over product characteristics, illustrated in the figure below by the blue contours. The manager
should also understand the distribution of of competitors' (and their own) products
on those same characteristics. These are illustrated as points in the figure below. A manager might then
decide to enter a market by offering a product with characteristics that customers value but where few
competing products exist. 

```{r, echo = F, warning = F, message = F}
library(ggplot2); library(dplyr)

tau <- c(1, 2)
Omega <- matrix(c(1, .5, .5, 1), 2,2 )

Sigma <- diag(tau) %*% Omega %*% diag(tau)

draws <- MASS::mvrnorm(5e4, c(1, .5), Sigma)

products <- MASS::mvrnorm(5, c(1, .5), Sigma)

ggplot(data = as.data.frame(draws), aes(V1, V2)) +
  geom_density_2d(aes(fill = ..level..), geom = "polygon", alpha = 0.2, colour = "blue") +
  labs(x = "Product characteristic 1",
       y = "Product characteristic 2",
       title = "Preference distribution and competing products") +
  geom_point(data = as.data.frame(products)) +
  ggthemes::theme_few() +
  annotate("text", x = 2, y = -3, label = "Existing products") +
  annotate("text", x = 0, y = 4.5, label = "Density of\ncustomer preferences", colour = "blue")
  
```

Such analysis might be quite straightforward when the manager has access to customer-transaction-level
data of their competitors, but this is not feasible in most cases. Instead, they need to make do 
with the sorts of aggregate data available from the same research houses in the example above.

## A generative model of consumer choice

These examples above have the same objective: we want to perform analysis that requires knowing about 
the distribution of customer preferences. And both problems face the same major constraint: we don't
observe transaction-level data for all products--we only observe aggregate sales in each market. 
Moreover, the most valuable aspects of consumer preferences are _relative_, not absolute - e.g. a 
manager would be interested not just in whether a particular product would sell, but _how well_ it 
would sell at different price points. How much would consumers be willing to pay for a new product, 
and how much would that affect the prices and sales of existing products? 

Even with very fine data and a few frequently purchased products, this is a difficult question to answer. 
It requires inferring how consumers would substitute one good for another across a range of prices for 
each good---the sheer number of combinations of possibilities that could be relevant is too high to 
even conceive of, let alone observe in data. Of course, we don't actually have to observe every possible 
combination of products and prices to have a predictive model of substitution patterns that performs 
pretty well. We just need a tractable, flexible generative model that can smooth out the space of 
possibilities in a sensible way.

One extremely flexible model is the aggregate _random coefficients logit_ model. In this model, customer $i$ in 
market $t$ has preferences over product $j$ such that their total utility is

$$
V_{ijt} = u_{ijt} + \epsilon_{ijt} \mbox{ where } \epsilon_{ij t} \sim \mbox{Gumbel}(.) 
$$
$$
u_{ijt} = \alpha_{i}p_{jt} + X_{jt}\beta_{i} + \xi_{jt} 
$$

This says that each consumer receives utility, $u_{ijt}$, from good $j$ according to its price, $p_{jt}$, its characteristics, $X_{jt}$, some systematic "demand" or "quality" shock applicable to all customers in 
market $t$, $\xi_{jt}$, and some iid shock, $\epsilon_{ijt}$. There are a couple of very important
things to notice about this utility function:

- We don't observe $u_{ijt}$ or $\xi_{jt}$. Because $\xi_{jt}$ is normally interpreted as "quality" or 
a "demand shock", we assume that it is correlated with price. 
- In this specification, each customer has their own preferences over price and product characteristics. The joint
distribution of $\alpha_{i}$ and $\beta_{i}$ **is** the distribution of _structural parameters_ we care about. 
It tells us what combinations of product characteristics and prices customers like, as well as how we expect 
their utility from a product to vary when the product characteristics or price change. 
- We make the assumption that $\epsilon_{ijt}$ is distributed according to a Gumbel distribution. This is 
a convenience that allows us to use a softmax function for the probabilities of purchase. The parameterization
of the Gumbel doesn't actually matter so long as all products, customers and markets have the same location
and scale parameters. Once we introduce the outside good below, we assume that the non-random portion of utility
for the outside good, $u_{i0t}$ is 0. 

### Introducing an outside good

In the model, a customer purchases whichever good gives them the highest utility. This means we only care about 
_relative_ utility, not absolute utility. We could add 4 or 200 to all products' utilities, and customer
choices would remain the same. This necessitates us introducing an "outside good" (with an expected utility of 0), 
which serves as a reference point against which all other utilities are measured. We make the 
assumption that it has sales equal to the potential size of the market (which we don't know) less the sales of 
the goods for which we have data. 

### From individual utilities to individual probabilities of purchase

Under the above assumptions, it is possible to derive the probability that an individual will purchase good $j$. 
This derivation is in Train (2003)---we'll spare you from it here. Individual purchase probabilities are just the 
probability that a good's associated utility is the highest of all available good (including the outside good).

$$
p(u_{ijt}=\max(u_{it})) = \frac{\exp(\alpha_{i}p_{jt} + X_{jt}\beta_{i} + \xi_{jt})}{1 + \sum_{j}\exp(\alpha_{i}p_{jt} + X_{jt}\beta_{i} + \xi_{jt})}
$$

Where the denominator includes the expected value of the utility of the outside good $1 = \exp(u_{0})$. 

### Generating aggregate sales data from the model

Because this is a generative model, there is a clear mapping between our economic model (above) and the observed
data. But first, what exactly is the observed data? We have the **inputs**, prices $p_{jt}$, and  $P$ observed product characteristics $X_{jt}$. Using these variables and the **unknowns**  $\alpha_{i}$, $\beta_{i}$, $\xi_{jt}$ fixed at
some values, we should be able to generate the **outcomes**, $y_{jt}$---the sales of each good in each market 
(some researchers use market shares instead of sales; we prefer sales as described below). 

But how do the data come about? If a customer's probability of choosing to purchase a given object is given by a combination of the product knowns and the unknowns of the model, then the market shares of each product will, for a large number of customers, be the integral of the individual probability model over the preference space. 
This sounds difficult, but is really quite simple: each customer has a probability of purchasing each good. If we 
have values for the unknowns of the model $\alpha_{i},\, \beta_{i},\, \xi_{jt}$ (and the model is correct) we 
know the probability of customer $i$ making each purchase. We can calculate market shares as the average of these
probabilities over many customers. Note that because the probability model is non-linear, this is different to 
the probabilities implied by the softmax function at the average values of $\alpha_{i},\, \beta_{i}$. 

What have we done? We've gone from a model of individual choice making to implied market shares of heterogeneous 
goods. This is often where researchers' models stop (in the famous BLP case, $\xi_{jt}$ is backed out to set
implied market shares to actual market shares). But we want to be able to deal with measurement error---the fact
that given some true market shares for each good $s_{jt}$, our data has been collected with error giving us
estimates $\hat{s}_{jt}$. We make the assumption that true market shares $s_{jt}$ map to observed sales (which
are observed with error) through a multinomial distribution with a given market size. 

$$
y_{jt} \sim \mbox{Multinomial}(s_{jt}, \mbox{Market size}_{t})
$$

### Modeling price

A major difference between our model and other approaches is to model the demand shock $\xi_{jt}$
as an explicit random variable, about which we want to perform inference. The approach we take is to model
it as a latent factor, whose value is revealed partly in utilities (and market shares), and partly
in prices or any other endogenous product characteristics (Petrin and Train (2010) propose a similar 
control function approach and show how it can substitute for IVs). This requires that we model the 
functional form of the true price-generating process. Here we'll make the assumption that the natural 
log of prices is a linear function of product characteristics and the latent factor. 

$$
\log(p_{jt}) \sim \mbox{Normal}(X_{jt}\gamma + Z_{jt}\delta +  \lambda \xi_{jt}, \sigma_{p})
$$
$\lambda$ is known as a _factor loading_. We've used a lognormal distribution here because it places no weight on 
negative prices. If negative prices are possible, we should not use this formulation. We could also use a less 
skewed distribution if desired, such as a truncated normal. $Z_{jt}$ is a vector of instruments for price (not included
in the utility model). By definition of $xi_{jt}$ as a random effect in this model implies orthogonality between 
the instruments and the structural shock; this is the same identifying assumption that justifies using GMM in
typical BLP.

Note that this need not be the full specification for price. Another, more economically valid approach similar to that 
in the original BLP paper would be to
model prices as being set by profit-maximizing competitors. That way, when we make predictions for sales of a new
product, those share predictions would be the sales _given that competitors will respond to the new product_. 

This structural add on is quite simple. If each product is sold by a separate firm, then firm $j$ faces a profit function

$$
\Pi_{jt} = (p_{jt} - mc_{jt})\times\text{market size}_{t}\times \text{share}_{j}(p_{t}, X_{t}, \xi_{t}, \alpha, \beta, L)
$$
where $mc_{jt}$ is the marginal cost of good j in market t (so $p_{jt} - mc_{jt} is the mark-up per product). 
Market size is the total number of goods sold (assumed to be 
independent of pricing choice), and $\text{share}_{j}(p_{t}, X_{t}, \xi_{t}, \alpha, \beta, L)$ the market share of good $j$, 
as a function of data and model parameters derived below. 

Firm $j$ maximizes profits, implying that pricing satisfies the first order condition

$$
\frac{\partial \Pi_{jt}}{\partial p_{jt}} = \text{share}_{j}(p_{t}, X_{t}, \xi_{t}, \alpha, \beta, L) +
(p_{jt} - mc_{jt})\times\frac{\partial \text{ share}_{j}(p_{t}, X_{t}, \xi_{t}, \alpha, \beta, L)}{\partial p_{jt}} = 0
$$
Which implies that in equilibrium

$$
p_{jt} = mc_{jt} -\frac{\text{share}_{j}(p_{t}, X_{t}, \xi_{t}, \alpha, \beta, L)}{\frac{\partial \text{ share}_{j}(p_{t}, X_{t}, \xi_{t}, \alpha, \beta, L)}{\partial p_{jt}}}

$$
If we assume that marginal cost of each good is modeled as a function of product attributes $X_{jt}$ and cost-shifting
instruments $Z_{jt}$, we could use a truncated normal model. 

$$
mc_{jt} \sim \text{Normal}_{+}(X_{jt}\psi + Z_{jt}\phi, \sigma_{mc}) \text{ or } mc_{jt} = X_{jt}\psi + Z_{jt}\phi + \eta_{jt} 
\text{ with } \eta \sim \text{Normal}_{>-(X_{jt}\psi + Z_{jt}\phi)}(0, \sigma_{mc})
$$

This implies a likelihood for price of

$$
p_{jt} = X_{jt}\psi + Z_{jt}\phi  - \frac{\text{share}_{j}(p_{t}, X_{t}, \xi_{t}, \alpha, \beta, L)}{\frac{\partial \text{ share}_{j}(p_{t}, X_{t}, \xi_{t}, \alpha, \beta, L)}{\partial p_{jt}}}+ \eta_{jt} 
\text{ with } \eta \sim \text{Normal}_{>-(X_{jt}\psi + Z_{jt}\phi)}(0, \sigma_{mc})
$$

Note this is still a latent factor model, but the impact of $\xi$ on price is now quite non-linear. 

### Estimating the model from aggregate market-level data 

At a high level, the trick to estimating this model is to estimate the the distribution of 
the individual-level coefficients, rather than the actual individual-level coefficients (which we obviously
cannot estimate from aggregate data). We do this by reformulating the utility model in terms of fixed
and random utility, and passing in the individual random effects $z_{i}$ as data.

First, express utility in terms of a fixed and random portion: 

$$
u_{ijt} = \alpha p_{jt} + X_{jt}\beta + (p_{jt},\, X_{jt}')\, L\, z_{i}' + \xi_{jt} 
$$

$z_{i}$ is a row vector of $P+1$ independent draws from some distribution, normally a unit normal, and $L$ is the lower triangular
Cholesky factorization of $\Sigma$, which is a $P+1 \times P+1$ covariance matrix. To be clear, $\Sigma$ is the covariance matrix of variations in $(\alpha_{i},\, \beta_{i}')$ across customers. If the element of $\Sigma$ corresponding to price and product characteristic 3 is negative, it 
means that customers who are more less sensitive to price (assuming all are negative, those whose $\alpha_{i}$s are closer to 0) 
tend to derive less utility from characteristic 3. Good estimates of $\Sigma$ are what give us good estimates of the 
distribution of preferences, which is what we ultimately want.

Note that if we have information about how markets differ from one another (for instance their demographics), 
we could include that information in the random effects part of the model. 

Given this structure, we can estimate the structural parameters (and demand shocks) using the following method: 

1. Draw a $NS \times P+1$ matrix independent shocks $z_{t}$, for some large number $NS$. We normally use the same shocks for every market. 
2. For a given draw of the structural parameters $\alpha,\, \beta,\, \xi_{jt}, \Sigma$, for each market for each $i\in 1:NS$ calculate $u_{ijt}$ and hence $p(u_{ijt}=\max(u_{it}))$.
3. Aggregate individual probabilities into predicted market shares $s_{jt,\mathrm{pred}}$
4. Model $y_{t}$ and $p_{t}$ as described above. 

Steps 2 and 3 occur in every iteration (or, if you are using HMC, every leapfrog step) of your model estimation. 

## Part 2: Fake data simulation

Astute readers will be aware that we always recommend simulating fake data with known parameters for a model. 
Here we do precisely that. All fake data simulation is in R. The comments should describe what's going on here. 

```{r}
set.seed(57)
# Dimensions of the data. 
NS <- 1000 # 1000 fake customers in each market
J <- 10 # 10 products
T <- 20 # 20 markets
P <- 3 # 3 covariates

# structural parameters
alpha <- -1
lambda <- .8
beta <- rnorm(P)

# Create a covariance matrix of the individual-level parameters

scales <- seq(from = .2, to = .9, length.out = P+1)

# Generate a random correlation matrix
XX <- matrix(rnorm(4*6), 6, 4)
Omega <- cor(XX)
Sigma <- diag(scales) %*% Omega %*% diag(scales)


# Product characteristics matrix
X <- matrix(rnorm(J*P), J, P)

# Easier to use if we repeat it for each market. We can have different 
# characteristics (like advertising) in each market
X_long <- do.call(rbind, replicate(T, X, simplify = F))

# structural shocks and price
xi <- rnorm(T*J, 0, 1)
xi_mat <- matrix(xi, T, J, byrow = T)
psi <- rnorm(P)

# The generative model for price
price <- exp(0 +  lambda*xi + X_long %*% psi + rnorm(T*J, 0, 0.5))
price_mat <- matrix(price, T, J, byrow = T)

# Market size
market_size <- round(rpois(T, 30000))

# Deltas (the fixed part of utility for each product)
delta <- alpha*price + X_long %*% beta + xi
delta_mat <- matrix(delta[,1], T, J, byrow = T)

# random shocks. (alpha_shocks, beta_shocks) = z_t
z <- matrix(rnorm(NS*(P+1)), NS, P+1)

# Empty market shares. Mat is for the observed products; sales is for all goods including the outside good
shares_mat <- matrix(NA, T, J)
sales <-  matrix(NA, T, J+1)

# Loop through each matrix and generate sales for each product
for(i in 1:T) {
  
  # Latent utility matrix
  utility <- matrix(NA, NS, J)
  # Create the random component of the (alpha, beta) vector
  random_effects <-  z %*% chol(Sigma)
  
  # monte carlo integration
  for(n in 1:NS){
    utility[n,] <-t( exp(delta_mat[i,] + cbind(price_mat[i,], X) %*% random_effects[n,]))
    utility[n,] <- utility[n,]/(1 + sum(utility[n,]))
  }
  shares_mat[i,] <- colMeans(utility)
  
  # Now we're going to observe the shares with measurement error
  # Last column is the outside good
  sales[i,] <- rmultinom(1, market_size[i], c(shares_mat[i,], 1 - sum( shares_mat[i,])))
}

```


It should be pointed out that here $\xi_{jt}$ and $p$ are correlated. This should introduce
endogeneity problems in the model. 

```{r, echo = F}
qplot(xi, log(price)) + ggthemes::theme_economist() +
  ggtitle("xi clearly related to price")
```


## Part 3: Writing out the model in Stan

Below we implement the model described above in Stan. 

A couple of things to look out for in the code: 

1. We pass $z_{t}$ in as two sets of shocks, one for $\alpha$ and one for $\beta$. There's no good reason for this. 
2. We stack $X$, a $J\times P$ characteristic matrix, $T$ times. In the DGP above, we assume that a product has
the same characteristics in each market. In reality, we would assume that things like advertising would vary across markets. 
3. Although we simulate the price above with instruments, below we don't use the instruments at all for estimation 
of the model. 

```
// our Stan model, saved as vsb.stan
// first we define the function that takes data and parameters and returns predicted market shares
functions {
  // calculates shares for a given market
  row_vector shares(real alpha, vector beta, matrix bigX, matrix Sigma, row_vector xi, matrix z) {
    matrix[rows(z), cols(xi)] utilities;
    matrix[rows(z), cols(xi)] probs;
    row_vector[cols(xi)] shares;
    // 1. Rather than pass in p and x separately, we'll pass in bigX = append_col(p, X)
    // 2. append alpha_shock, beta_shock
    {
      matrix[rows(z), cols(xi)] tmp;
      
      tmp = rep_matrix((bigX*append_row(alpha, beta) + xi')', rows(z));
      
      // replace the append_col wing single values (might speed things up)
      utilities = exp(tmp + z * cholesky_decompose(Sigma)' * bigX');
      
      for(i in 1:rows(z)) {
         probs[i] = utilities[i]/(1 + sum(utilities[i]));
      }
      
    }
    
    for(j in 1:cols(probs)) {
      shares[j] = mean(col(probs, j));
    }
    
    return(shares);
  }
}
// next define our data
data {
  int NS; // number of individuals in integration
  int J; // number of products
  int T; // number of markets
  int P; // number of features
  matrix[NS, P+1] z; // normal(0,1) draws of the shocks
  matrix[T, J] price; // price for each unit
  int sales[T, J+1]; // unit sales across T markets for J+1 products (inc outside good)
  matrix[T*J, P] X_repeat; // T Xs stacked on top of each other. This format allows characteristics to vary across markets.
  real nu;
}
// next join the product data together into single matrices
transformed data {
  matrix[T*J, P+1] bigX;
  bigX = append_col(to_vector(price'), X_repeat);
  
}
// define parameters
parameters {
  real alpha; 
  vector[P] beta;
  vector[P] gamma;
  real gamma0;
  real<lower = 0> price_scale;
  matrix[T, J] xi;
  vector<lower = 0>[P+1] scales;
  corr_matrix[P+1] Omega;
  real lambda;
}

transformed parameters {
  cov_matrix[P+1] Sigma;
  Sigma = quad_form_diag(Omega, scales);
}
// and the model
model {
  // priors
  alpha ~ normal(0, 1);
  beta ~ normal(0, 1);
  gamma0 ~ normal(0, 1);
  gamma ~ normal(0, 1);
  price_scale ~ normal(0, 1);
  lambda ~ normal(0, 1);
  to_vector(xi) ~ normal(0, 1);
  scales ~ inv_gamma(2,1);
  Omega ~ lkj_corr(nu);
  
  // model of price -- this helps pin down xi
  to_vector(price') ~ lognormal(gamma0 + X_repeat * gamma + lambda * to_vector(xi'), price_scale);
  
  // model of sales 
  {
    matrix[T, J+1] pred_shares;
    for(t in 1:T) {
      // predicted market shares given data and parameters
      pred_shares[t,1:J] = shares(alpha, beta, bigX[(t*J-J+1):(t*J)], Sigma, xi[t], z);
      pred_shares[t,J+1] = 1 - sum(pred_shares[t,1:J]);
      // sales are measured with multinomial measurement error
      sales[t] ~ multinomial(pred_shares[t]');
    }
    
  }
}
```

```{r, results = "hide"}

# run model ---------------------------------------------------------------

# Compile Stan function to check that it generates sensible shares

library(rstan)
options(mc.cores = parallel::detectCores())

data_list <- list(NS = NS, 
                  J = J, 
                  T = T, 
                  P = P, 
                  z = z, 
                  price = price_mat, 
                  sales = sales,
                  X_repeat = X_long,
                  mkt_sizes = as.vector(market_size),
                  nu = 4)

# Compile the model
compiled_model <- stan_model("vsb.stan")

# For the sake of time, we estimate this using optimization
test_optim <- optimizing(compiled_model, data = data_list)
  
```


Now how did the model go at recapturing known demand shocks? 

```{r, echo = F}
library(dplyr); library(ggplot2); library(ggthemes)

xi_est <- test_optim$par[grepl("xi", names(test_optim$par))]
xi_est <- as.vector(t(matrix(xi_est, T, J)))

data_frame(Estimates = xi_est, 
           `True values` = xi) %>% 
  ggplot(aes(x = `True values`, y = Estimates)) + 
  geom_point() + 
  geom_abline(aes(intercept = 0, slope = 1)) +
  theme_economist() +
  labs(title = "Estimates and true values of xi")

```

And how about the structural parameters? 

```{r}
Omegas <- test_optim$par[grepl("Omega", names(test_optim$par))]
Scales <- test_optim$par[grepl("^scale", names(test_optim$par))]
pars <- c(test_optim$par[1:(P+1)], Omegas, Scales)
true_values <- c(alpha, beta, as.vector(Omega), scales)

data_frame(Estimates = pars, 
           `True values` = true_values,
           Parameter = c("alpha", rep("beta", P), rep("Omega", (P+1)^2), rep("scale", P+1) )) %>% 
  ggplot(aes(x = `True values`, y = Estimates)) + 
  geom_point(aes(colour = Parameter)) + 
  geom_abline(aes(intercept = 0, slope = 1)) +
  theme_economist() +
  labs(title = "Estimates and true values\nof structural parameters")
```

**_Voila!_**. That ran in about 25 seconds. 

## Conclusion

What we've done above is shown that it can be extremely simple to express a random coefficients logit model
with measurement error using a latent factor approach, and importantly, without any instruments. This
is a pretty radical departure from the most popular methods in the literature. 

The other big advantage is that, given you have enough computation time, the model can be estimated using 
Bayesian techniques. In the code above, we could do this by using `sampling()` rather than `optimizing`. 
With the current code, this is quite slow, but does generate robust parameter estimates. If your decision
problem requires genuine estimates of uncertainty around the structural parameters, I'd recommend experimenting
using optimization, and then using HMC (`sampling`) for the production run of the model. In my experiments, 
I've found they generate similar point-estimates. 

If you've any comments, please get in touch. 

