---
title: "Does the New York City Police Department Rely on Quotas?"
author: | 
  | Jonathan Auerbach
  | Department of Statistics
  | Columbia University
date: "12/17/2017"
output: pdf_document
bibliography: auerbach_jonathan.bib
---

\vfill

##Abstract

We investigate whether the New York City Police Department (NYPD) uses 
productivity targets or quotas to manage officers in contravention of New York 
State Law. The analysis is presented in three parts. First, we introduce the 
NYPD's employee evaluation system, and summarize the criticism that it 
constitutes a quota. Second, we describe a publically available dataset of 
traffic tickets issued by NYPD officers in 2014 and 2015. Finally, we propose a 
generative model to describe how officers write traffic tickets. The fitted 
model is consistent with the criticism that police officers substantially alter 
their ticket writing to coincide with departmental targets. We conclude by 
discussing the implications of these findings and offer directions for further 
research.

\newpage
  
## I. Introduction

Critics have periodically accused the New York City Police Department (NYPD) of 
relying on traffic ticket quotas to evaluate police officers and maintain 
productivity [@bacon2009bad, pg. 97], [@rayman2013nypd, pg. 49, pg. 64], 
[@eterno2012crime, pg. 170]. Such concerns are far from frivolous. Traffic 
ticket quotas are illegal in New York. Section 215-a of New York State Labor Law 
specifically prohibits NYPD supervisors from ordering an officer to write a 
predetermined number of tickets over a predetermined period of time.

Despite a lengthy history [@burrows1998gotham], [@wallace2017greater], 
accusations of quotas came to a head only recently, when the commander of the 
75th precinct issued a memo directing officers to write ten tickets a month. The 
officers' union filed a grievance, and in 2006 an arbitrator ruled in their 
favor, ordering that "The city shall cease and desist from maintaining a 
vehicular ticket quota." [@fahim2006] Yet instances of explicit ticket quotas 
continue to occur. The most notable came in 2010 when Officer Schoolcraft, a 
whistleblower from the 81st precinct, recorded his supervisor directing officers 
to write five traffic tickets a week each for drivers not wearing a seat belt, 
using a cellphone, double-parking and parking in a bus stop.

While the NYPD cannot legally set quotas, the law does not forbid the 
consideration of past productivity when assessing officers. In fact, the NYPD's
evaluation system --- Operations Order number 52, series 2011, titled 
"Quest for Excellence - Police Officer Performance Objectives" (hereafter 
abbreviated Q4E) --- mandates it. Under Q4E, supervisors are directed to assess 
officers on the 7th, 14th and 21st of each month according to their 
productivity. At the end of the month, supervisors rate each officer as 
effective or ineffective. These supervisors are then rated by their own managers 
at the end of each quarter. 

The pressure to do more with fewer resources leaves supervisors walking the fine 
line between incentivizing officer productivity and ordering it as 
Schoolcraft's supervisor did. The line is blurred when supervisors base 
evaluations on an officer's productivity relative to their peers. Spurring low 
performing officers to catch up to peers induces a "ticket race" between 
officers. Since supervisors control how officers are compared to the group, they 
control the pace of the race. The faster the race, the greater the number of 
tickets expected of each officer each period, effecting a de facto quota system 
[@dornsife1978ticket, pg. 3].

The legality of this practice has yet to be determined [@scheidlin2013], but
pressuring officers to increase productivity to compensate for past behavior 
reduces discretion and violates the spirit of the law as stated in the 1978 
memorandum accompanying the original anti-quota bill:

> The police officer as well as the public, need not be put under the pressure
> of a mandatory ticket quota. Such a policy can only hurt the effectiveness of 
> the police officer in the performance of his other duties while he must give
> the driving public a rash of summonses to meet the quota.

The ticket race is a plausible mechanism by which NYPD supervisors circumvent 
the anti-quota law and compel officers to give "a rash of summonses". However, 
the vast majority of evidence implicates only specific supervisors over short 
periods of time, without quantifying the magnitude or scope of the practice. 
Fortunately, traffic ticket data can be used determine if quota-like behavior 
exists across multiple precincts and time periods, or whether productivity 
targets reflect isolated incidents, confined to the times and places where 
physical evidence has been made available. The goal of the present analysis is 
to demonstrate how such an analysis might be performed.

## II. Data

The primary dataset contains every recorded summons issued to drivers for a 
parking or moving violation in New York City between January 2014 and December 
2015. It includes the id number of the issuing officer, their precinct, the date 
the ticket was written and the violation for which the ticket was written. The 
first five observations of the dataset are displayed in Table 1. Summary 
statistics for the first three columns are displayed in Table 2, and the most 
common traffic ticket violations are displayed in Table 3. Note that the ticket 
types from Officer Schoolcraft's recordings rank at 1, 4, 5 and 8 among the top 
ten ticket types.

```{r setup, message = FALSE, warning = FALSE}
library("knitr")
library("timeDate")
library("scales")
library("tidyverse")
library("lubridate")
library("stringr")
library("rstan")
library("splines")

theme_set(theme_bw())
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

download.file(str_c("https://raw.githubusercontent.com/jauerbach/",
                "Seventy-Seven-Precincts/master/data/tickets.csv.zip"), 
              str_c(getwd(),"/tickets.csv.zip"))
tickets <- read_csv("tickets.csv.zip", col_types = cols(id = col_double())) 

crashes <- 
  read_csv(str_c("https://raw.githubusercontent.com/jauerbach/",
                 "Seventy-Seven-Precincts/master/data/crashes.csv"))
```

```{r summary, message = FALSE, warning = FALSE}
tickets %>% 
  head() %>% 
  mutate(violation = substr(violation, 1, 48)) %>%
  kable(caption="Example Observations of Dataset")

tickets %>% 
  select(-violation) %>%
  summary() %>% 
  kable(caption="Summary of Dataset")

tickets %>% 
  count(violation) %>% 
  arrange(desc(n)) %>%
  mutate(violation = substr(violation, 1, 48)) %>%
  head(n = 10) %>%
  kable(caption= "Most Common Traffic Tickets")
```

The unit of this analysis is the number of tickets written by each officer each 
day, and the following code block aggregates observations by officer and day. 
Days of the month are grouped into four review periods depending on their 
proximity to the Q4E evaluations on the 7th, 14th, 21st and end of the month. 
For example, the first review period contains the first seven days of the month, 
the second review period contains the eighth of the month to the fourteenth, and 
so on.

Some officers are only represented for a portion of the two-year sample period.
We assume an officer is eligible to write tickets for every day in a month if 
at least one ticket was written by the officer that month. Otherwise, the 
officer is excluded for the entire month. The only exception is days during 
which no ticket is written in the entire precinct. Those days are excluded from
the analysis.

```{r data, message = FALSE, warning = FALSE}
# calculate the number of tickets per day. 
by_day <- tickets %>% 
  mutate(month_year = format(date, "%m-%Y")) %>%
  group_by(id, command, date, month_year) %>%
  summarise(daily_tickets = n()) %>%
  ungroup()

# add zeros and period mean and max, assuming at least one ticket written in 
## command during month for a day to be eligible for writing tickets
by_day <- by_day %>%
  group_by(command, month_year) %>%
  nest() %>% 
  mutate(elig_days = map(data, function(df) df %>% expand(id, date)),
         data_aug = map2(elig_days, data, left_join, by = c("id","date"))) %>%
  select(command, month_year, data_aug) %>%
  unnest() %>% 
  mutate(daily_tickets = ifelse(is.na(daily_tickets), 0, daily_tickets))

# add mean tickets and max tickets for each command each review period. review 
## periods are on the 7th, 14th, 21st and end of each month
by_day <- by_day %>%
  mutate(period = cut(parse_number(format(date, "%d")),
                      breaks = c(1,7,14,21,31), labels = FALSE,
                      right = TRUE, include.lowest = TRUE)) %>%
  group_by(command, month_year, period) %>%
  nest() %>% 
  mutate(mean = map_dbl(data, function(df) df %>%
                          group_by(id) %>%
                          summarise(sum(daily_tickets)) %>% 
                          pull %>%
                          mean()),
         median = map_dbl(data, function(df) df %>%
                          group_by(id) %>%
                          summarise(sum(daily_tickets)) %>% 
                          pull %>%
                          median()),
         max = map_dbl(data, function(df) df %>%
                         group_by(id) %>%
                         summarise(sum(daily_tickets)) %>% 
                         pull %>%
                         max())) %>%
  unnest()

#add number of tickets per officer per period
by_day <- by_day %>% 
  left_join(by_day %>%
              group_by(command, month_year, period, id) %>%
              summarise(period_tickets = sum(daily_tickets)) %>%
              ungroup(),
            by = c("command", "month_year", "period", "id"))

#add number of tickets, command max/mean in pervious period
by_day <- by_day %>% 
  left_join(by_day %>%
              select(-daily_tickets, - date) %>%
              unique() %>%
              mutate(period = period + 1) %>%
              rename(mean_prev = mean,
                     median_prev = median,
                     max_prev = max,
                     tickets_prev = period_tickets),
            by = c("command", "month_year", "period", "id")) %>%
  mutate(mean_prev = ifelse(is.na(mean_prev), 0, mean_prev),
         median_prev = ifelse(is.na(median_prev), 0, mean_prev),
         max_prev = ifelse(is.na(max_prev), 0, max_prev),
         tickets_prev = ifelse(is.na(tickets_prev), 0, tickets_prev))

#add US Holidays
by_day <- by_day %>% 
  left_join(sapply(c("USNewYearsDay", "USMLKingsBirthday", "USWashingtonsBirthday", 
                     "USMemorialDay", "USIndependenceDay", "USLaborDay",
                     "USColumbusDay", "USElectionDay", "USVeteransDay",
                     "USThanksgivingDay", "USChristmasDay"), 
                   function(x) as.Date(holiday(2014:2015, x))) %>%
              as_tibble() %>%
              gather(key = "holidays", value = "date") %>%
              mutate(date = as.Date(date, origin = "1970-01-01")),
            by = "date") %>% 
  mutate(holidays = ifelse(is.na(holidays), "USNone", holidays))
```

## III. Preliminary Evidence

For computational reasons, we limit the present analysis to Schoolcraft's 
Precinct, Precinct 81.

The data indicate that more tickets are written in the second half of the month,
and this increase comes from officers that are behind their peers. Figure 1.1 
demonstrates that the average number of tickets written per day increases from 
the first half to the second half of the month. Here, the second half of the 
month begins on the 15th, following the second evaluation date under Q4E. The 
increase between the two halves is six percent.

```{r evidence1, message = FALSE, warning = FALSE}
by_day %>% 
  filter(command == 81) %>%
  mutate(half = ifelse(period > 2, 
                       "Second Half \nof Month", 
                       "First Half \nof Month")) %>%
  group_by(half) %>%
  summarise(y = sum(daily_tickets)) %>%
  mutate(y = ifelse(half == "First Half \nof Month", 
                    y / (24 * 14), 
                    y * 12 / (24 * 198))) %>%
  ggplot() +
    aes(half, weight = y) +
    geom_bar() +
    theme(legend.position = "bottom") +
    scale_y_continuous(labels = comma) +
    labs(y = "average number of tickets written in 2014-15", x = "", 
         fill = "",
         title="Figure 1.1: 
         the ticketing rate increases six percent over the month ...")
```

Figure 1.2 demonstrates that the increase in Figure 1.1 is unlikely the result 
of driver behavior since the average number of vehicle crashes does not increase 
over this period. If changes in exposure were causing the increase in Figure 
1.1, more crashes would be expected as well.

```{r evidence2, message = FALSE, warning = FALSE}
crashes %>% 
  filter(Precinct == 81) %>%
  mutate(period = cut(parse_number(format(DATE, "%d")),
                      breaks = c(1,7,14,21,31), labels = FALSE,
                      right = TRUE, include.lowest = TRUE),
         half = ifelse(period > 2, 
                       "Second Half \nof Month", 
                       "First Half \nof Month")) %>%
  group_by(half) %>%
  summarise(y = sum(n)) %>%
  mutate(y = ifelse(half == "First Half \nof Month",  
                    y / (24 * 14), 
                    y * 12 / (24 * 198))) %>%
  ggplot() +
    aes(half, weight = y) +
    geom_bar() +
    theme(legend.position = "bottom") +
    scale_y_continuous(labels = comma) +
    labs(y = "average number of collisions in 2014-15", x = "", 
         fill = "",
         title="Figure 1.2: 
         ...while the number of collisions remains unchanged")
```

Figure 1.3 shows that the increase in the number of tickets issued is due 
entirely to officers who have below median productivity. Since there are more 
days after the 15th than before it, a constant number of traffic ticket 
translates to a decrease in the average number of tickets written. Much of that 
decrease occurs in the fourth period as displayed in Figure 1.4.

```{r evidence3, message = FALSE, warning = FALSE}
by_day %>% 
  filter(command == 81) %>%
  mutate(prod = ifelse(tickets_prev >= median_prev, 
                       "Officer >= Median","Officer < Median"),
         half = ifelse(period > 2, 
                       "Second Half \nof Month", 
                       "First Half \nof Month")) %>%
  group_by(half, prod) %>%
  summarise(y = sum(daily_tickets)) %>%
  ggplot() +
    aes(half, weight = y, fill = factor(prod)) +
    geom_bar() +
    theme(legend.position = "bottom") +
    scale_y_continuous(labels = comma) +
    labs(y = "all traffic tickets written in 2014-15", x = "period", fill = "",
         title="Figure 1.3: 
         officers behind their peers account for the entire increase in the 
         number of tickets written in the second half of the month ...")

by_day %>% 
  filter(command == 81,
         period != 1) %>%
  mutate(prod = ifelse(tickets_prev > median_prev, 
                       "Officer >= Median","Officer < Median")) %>%
  group_by(period, prod) %>%
  summarise(y = sum(daily_tickets)) %>%
  mutate(y = ifelse(period == 4, 
                    y * 12 / (24 * 134), 
                    y / (24 * 7))) %>%
  ggplot() +
    aes(period, weight = y, fill = factor(prod)) +
    geom_bar(position = "dodge") +
    theme(legend.position = "bottom") +
    scale_y_continuous(labels = comma) +
    labs(y = "average number of tickets written in 2014-15", 
         x = "period", fill = "", 
         title="Figure 1.4: 
         ... but it is officers ahead of their peers that are drastically 
         reducing their ticket writing rate")
```

The data are consistent with the criticism that officers winning the ticket 
race reduce the number of tickets written each period and officers losing the 
ticket race increase the number of tickets [@dornsife1978ticket, pg. 3]. The 
problem with the analysis so far is that these changes could be a selection 
effect. If officer activity were random each period, one would expect an 
increase after selecting the lowest performing officers and a decrease after 
selecting the highest performing officers. Productivity should regress towards 
the mean in either case by default, as an artifact of the selection process.

A model is needed to fully evaluate the working theory. However, any model is 
complicated by the fact that the majority of officers wrote a minority of 
traffic tickets. As can be seen in the Figure 1.5, over the two-year study 
period, ninety percent of officers wrote one or fewer tickets each day, while 
the top one percent of officers wrote more than one thousand tickets. However, 
these officers are responsible for the increase in the overall ticket rate and 
their behavior, not the behavior of the most common ticket writers, is of 
primary interest.

In this sense, the data is sparse, and information is spread out among a large 
number of relatively rare events. Determining whether officers have increased 
their ticketing rate fits well within the Bayesian hierarchical model paradigm 
for this reason: information can be pooled across officers to make generalizing 
statements of officer behavior.

```{r evidence4, message = FALSE, warning = FALSE}
by_day %>% 
  filter(command == 81) %>%
  ggplot() +
    aes(daily_tickets) +
    geom_histogram() +
    scale_y_log10(labels = comma) +
    labs(x = "number of tickets written per officer per day in 2014-15", 
         y = expression(log[10] ~ " number of officers"),
         title="Figure 1.5: 
         the rate officers write tickets has a heavy tail")
```

## IV. Model

We begin by combining two log-linear models for contingency tables. Let $y_{od}$ 
be the number of traffic tickets written by the $o^{th}$ officer on the $d^{th}$ 
day, and let $z_{d}$ be the number of crashes in the officers' precinct on the 
$d^{th}$ day. We include an effect for each day of the week, $w$, holiday, $h$, 
period, $p$, and month $m$. Week, holiday, period and month effects reflect 
various administrative policies. These effects are then pooled as follows:

\begin{align*} 
\sigma_{\cdot} &\sim \text{ Chi-Squared}(1) \\ 
\beta_{\cdot} &\sim \text{Normal}(0, \sigma_{\cdot}) \\ 
\epsilon_d &\sim \text{Normal}(\beta_0, 1) \\
y_{od} &\sim \text{Poisson}( \text{exp} \{ \beta_o + \beta_w + 
    \beta_p + \beta_h + \beta_m + \sigma_{\epsilon} \epsilon_d \} ) \\
z_{d} &\sim \text{Poisson}( \text{exp} \{ \beta_0 + \sigma_0 \epsilon_d \} ) \\    
\end{align*}

The choice of $\text{Chi-Squared}(1)$ priors was arbitrary. The model was also 
fit with $\text{Normal}(0, 1)$ priors on the standard deviation parameters, and 
little change was observed in the outcomes.

```{r model1, message = FALSE, warning = FALSE}
model_data <- by_day %>% 
   filter(command ==  81) %>%
  mutate(id2 = as.numeric(as.factor(id)),
         date2 = as.numeric(as.factor(date)),
         months = months(date),
         months2 = as.numeric(format(date, "%m")),
         weekdays = weekdays(date),
         weekdays2 = wday(date),
         holidays2 = as.numeric(as.factor(holidays))) %>%
  with(list(N = nrow(by_day %>% filter(command ==  81)),
            O = length(unique(id2)),
            W = length(unique(weekdays2)),
            M = length(unique(months2)),
            H = length(unique(holidays2)),
            P = length(unique(period)),
            officer = id2,
            week = weekdays2,
            month= as.numeric(months2),
            holiday= as.numeric(holidays2),
            period = as.numeric(period),
            day = date2,
            y = daily_tickets,
            D = crashes %>% 
                filter(Precinct == 81) %>% 
                semi_join(by_day %>% filter(command ==  81), 
                          by = c("DATE" = "date")) %>% 
                  select(n) %>% 
                  pull() %>% 
                  length() ,
            z = crashes %>% 
                filter(Precinct == 81) %>% 
                semi_join(by_day %>% filter(command ==  81), 
                          by = c("DATE" = "date")) %>% 
                  select(n) %>% 
                  pull(),
            X = ifelse(tickets_prev - median_prev > 10, 10,
                       tickets_prev - median_prev)))

fit1 <- sampling(stan_model(file = "model1.stan"), 
                 data = model_data, 
                 iter = 500, 
                 control = list(max_treedepth = 15, adapt_delta = 0.9))
fit1_summary <- summary(fit1)
```

We use RStan [@RStan] to run four chains for 500 iterations each that 
approximate samples from the posterior distribution of the above model. The 
first 250 iterations are discarded in warm-up, leaving one thousand posterior 
draws to estimate the parameters. After increasing the max_treedepth and 
adapt_delta arguments, no notable issues occurred during sample. An overview of 
parameter estimates is provided in Figure 2, in which the average ticket writing 
of officers in Precinct 81 is decomposed into baseline, week, period and month 
effects (holiday effects not shown). Period and month patterns are evident, 
showing an increase before evaluations at the end of the month and quarter 
respectively. We also observe the decline around Christmas of 2014, which was 
the result of an alleged slowdown.

```{r model1 effect plots, warning = FALSE, message = FALSE}
var <- rownames(fit1_summary[[1]]); est <- fit1_summary[[1]][,1]
dfs <- list(
	date = by_day %>%
	        filter(command ==  81) %>% 
	        transmute(date, 
	                  per = period,  
	                  week_day = weekdays(date), 
	                  mnth = month(date), 
	                  daily_tickets) %>% 
					 group_by(date, per, week_day, mnth) %>% 
	         summarise(`log tickets` = log(mean(daily_tickets))),
	base = crashes %>% 
	          filter(Precinct == 81) %>% 
	          semi_join(by_day %>% 
	          filter(command ==  81), by = c("DATE" = "date")) %>% 
	          transmute(date = DATE, 
	                    baseline = est[substr(var,1,7) == "sigma_0"] * 
	                                  est[substr(var,1,6) == "epsilo"]
	                    ),
	day = tibble(`week day` = est[substr(var,1,7) == "sigma_2"] * 
	               est[substr(var,1,6) == "beta_2"], 
	             week_day = c("Sunday", "Monday", "Tuesday", "Wednesday", 
	                          "Thursday", "Friday", "Saturday")),
	period = tibble(period = est[substr(var,1,7) == "sigma_5"] * 
	                  est[substr(var,1,6) == "beta_5"], per = 1:4),
	month = tibble(month = est[substr(var,1,7) == "sigma_3"] * 
	                 est[substr(var,1,6) == "beta_3"], mnth = 1:12))

reduce(dfs, left_join) %>% 
	gather("effect", "value", `log tickets`, baseline, `week day`, period, month) %>%
	ggplot() +
		aes(date, value, color = effect) +
		geom_line() +
		facet_grid(factor(effect, 
		  levels = c("log tickets", "baseline", "week day", "period", "month")) ~ ., 
		  scales = "free") +
		theme_bw() +
		theme(legend.position = "none") +
		scale_x_date(date_labels = "%b", 
		      breaks = parse_date(c("2014-01-01","2014-04-01","2014-07-01",
		                            "2014-10-01", "2015-01-01","2015-04-01",
		                             "2015-07-01","2015-10-01", "2016-01-01"))) +
		     labs(x = "", y = "",
         title="Figure 2: 
         decomposition of log ticket rate implied by model")
```

Uncertainty intervals for the effects [@gelman2005analysis], [@gelman2006data] 
are depicted in Figures 3.2 through 3.4. Fat black lines correspond to 50 
percent uncertainty intervals, while thin black lines correspond to 95 percent 
intervals. We interpret these intervals as depicting plausible and possible 
values of the parameters.

Weekends and Wednesdays have lower ticket writing rates even after adjusting for 
the number of crashes. This likely corresponds with alternate side parking rules 
that disproportionately effect Monday/Wednesday and Tuesday/Fridays. March, June 
and September have higher ticket writing rates, which makes sense because, at 
the end of these months, supervisors are themselves evaluated. July and December 
have lower ticket writing rates. These are the traditional vacation months. 
November may be higher because of a pre-vacation push before the end of the 
year.

Periods are not statistically different from each other after adjusting for the 
other covariates. This is not necessarily inconsistent with quota-like behavior
since we would expect officers ahead in the ticket race to decrease their 
activity while officers behind would increase it. This reflection motivates the 
following refinement to the model.

```{r model1 coef plots, warning = FALSE, message = FALSE}
plot_coef <- function(summary, coef, label = NULL, str = 6) {
  if(is.null(label)) {
  var_name = rownames(summary$summary)[substr(rownames(summary$summary),1,str) == coef]
  } else {
    var_name = label
  }
  data.frame(
    name = var_name,
    mean = summary$summary[substr(rownames(summary$summary),1,str) == coef, "mean"],
    lower1 = summary$summary[substr(rownames(summary$summary),1,str) == coef, "mean"] -
      summary$summary[substr(rownames(summary$summary),1,str) == coef, "sd"],
    upper1 = summary$summary[substr(rownames(summary$summary),1,str) == coef, "mean"] +
      summary$summary[substr(rownames(summary$summary),1,str) == coef, "sd"],
    lower2 = summary$summary[substr(rownames(summary$summary),1,str) == coef, "mean"] -
      2 * summary$summary[substr(rownames(summary$summary),1,str) == coef, "sd"],
    upper2 = summary$summary[substr(rownames(summary$summary),1,str) == coef, "mean"] +
      2 * summary$summary[substr(rownames(summary$summary),1,str) == coef, "sd"]) %>%
    ggplot() +
      aes(x = name, y = exp(mean), ymin = exp(lower1), ymax = exp(upper1)) +
    geom_linerange(size = 2) +
    geom_linerange(aes(ymin = exp(lower2), ymax = exp(upper2))) +
    coord_flip()
}

qplot(fit1_summary$summary[
  substr(rownames(fit1_summary$summary),1,6) == "beta_1", "mean"]) +
  labs(x = expression(e^beta[officer]), y = "",
         title="Figure 3.1: 
         distribution of officer effects in Precinct 81")

plot_coef(summary = fit1_summary, coef = "beta_2",
          label = factor(c("Sunday", "Monday", "Tuesday", "Wednesday",
                           "Thursday", "Friday", "Saturday"), 
                         levels = weekdays(x=as.Date(seq(7), 
                                                     origin="1950-01-01")))) + 
  labs(x ="", y = expression(e^beta[w]),
         title="Figure 3.2: 
         day of week effects in Precinct 81")

plot_coef(summary = fit1_summary, coef = "beta_5", label = 1:4) +
  labs(x ="period effect", y = expression(e^beta[p]),
       title="Figure 3.3: 
         period effects in Precinct 81")

plot_coef(summary = fit1_summary, coef = "beta_4",
          label =   factor(c("Columbus Day", "Election Day", 
                             "Independence Day", "Labor Day", 
                             "Memorial Day", " M L Kings Birthday", 
                             "New Years Day", "None", "Thanksgiving Day", 
                             "Veterans Day", "Washingtons Birthday")) %>%
    (function(x) factor(x, levels = x[c(7, 6, 11, 5, 3, 4, 1, 10, 2, 9, 8)]))) +
  labs(x ="", y = expression(e^beta[h]),
  title="Figure 3.4: 
         holiday effects in Precinct 81")

plot_coef(summary = fit1_summary, coef = "beta_3",
          label = month(x=as.Date(29 * seq(12), origin="1950-01-01"), 
                        label = TRUE)) +
  labs(x ="", y = expression(e^beta[m]),
        title="Figure 3.5: 
         month effects in Precinct 81")
```

We now replace the period effect, $\beta_p$, with $f(rp_{o,p-1})$ where 
$rp_{o,p-1}$ is the relative productivity of officer $o$ in the previous period 
$p-1$. Relative productivity is defined to be the number of tickets written by 
the officer in the previous period minus the median number of tickets written by 
all officers of that period in the same precinct. When $p = 1$, $rp$ is taken to 
be zero for all officers. $f$ is assumed to be smooth and approximated by a 
penalized B-spline, with code appropriated, with minor changes, from @milad2017. 
For identifiability reasons, officers more than 10 tickets ahead are considered 
10 tickets ahead.

\begin{align*} 
\sigma_{\cdot} &\sim \text{ Chi-Squared}(1) \\ 
\beta_{\cdot} &\sim \text{Normal}(0, \sigma_{\cdot}) \\ 
\epsilon_d &\sim \text{Normal}(\beta_0, 1) \\
y_{od} &\sim \text{Poisson}( \text{exp} \{ \beta_o + \beta_w + 
    f(rp_{o,p-1}) + \beta_h + \beta_m + \sigma_{\epsilon} \epsilon_d \} ) \\
z_{d} &\sim \text{Poisson}( \text{exp} \{ \beta_0 + \sigma_0 \epsilon_d \} ) \\    
\end{align*}

A plot of $f$ is displayed below with standard error lines. The shape of $f$ is 
quite dramatic, suggesting that past ticket writing behavior drives future 
behavior. If an officer is five tickets below the median, their ticketing rate 
is expected to be ten percent above the average in the following period. If an 
officer is at the median, their ticketing rate is expected to be twenty percent 
below the average. Finally, if the officer is five tickets above the median, the 
ticketing rate is thirty percent below the average.

``` {r model 2, message = FALSE, warning = FALSE}
knots <- c(-5, -3, -1, 0, 1, 3, 5, 7, 10)
num_knots <- length(knots)
spline_degree <- 3

fit2 <- sampling(stan_model(file = "model2.stan"), data = model_data,
                 chains = 4, iter = 500,
                 control = list(max_treedepth = 15, adapt_delta = 0.9))
fit2_summary <- summary(fit2)
```

``` {r model2 coef plots, message = FALSE, warning = FALSE}
f_coef <- fit2_summary$summary[substr(rownames(fit2_summary$summary),1,6) == "lambda", 1]
f_sd <- fit2_summary$summary[substr(rownames(fit2_summary$summary),1,6) == "lambda", 3]

qplot(model_data$X, exp(f_coef), geom= "line") + 
  geom_line(aes(x, exp(y)), 
            data = data.frame(x = model_data$X, y = f_coef + f_sd),
            linetype = 2) +
  geom_line(aes(x, exp(y)), 
            data = data.frame(x = model_data$X, y = f_coef - f_sd),
            linetype = 2) +
  xlim(-4.5, 4.5) +
  labs(x = expression(rp["o,p-1"]), y = expression(e^f(rp["o,p-1"])),
       title="Figure 4: 
         effect of position in ticket race in Precinct 81")
```

## V. Conclusion

The fitted model is consistent with the criticism that police officers alter 
their ticket writing to coincide with departmental targets. Figure 4 indicates
that the behavior of Precinct 81 officers in past periods influences their 
future behavior, and that this influence is quite dramatic, accounting for as 
much as a fifty percentage point spread in the rate officers write tickets. Were 
the ticketing rate solely a function of road conditions, we would expect past 
behavior to have no effect on future behavior, and little change should be 
observed in Figures 1 or 4.

We conclude by making several, distinct remarks. We first point out that other 
theories could explain the pattern observed in the data. For example, the 
relationship depicted in Figure 4 could be a consequence of how the NYPD deploys 
officers. Rotating schedules each period would result in low ticketing officers 
one period increasing their productivity in the next. This seems unlikely, 
however, since nothing in Q4E or the literature suggests that scheduling takes 
place in a manner that regularly coincides with Q4E review dates. Nevertheless, 
we could account for this explanation by expanding the model to include 
interactions between officer and day of week or period effects. In addition, 
pooling across multiple precincts would also provide evidence that the estimated 
relationship is not due to precinct specific practices.

Regardless, the increases observed in Figures 1 and 4 are evidence of increased 
officer activity. We also point out that this increase is barely perceptible to 
the average driver. Even the largest estimated increase, a twenty-five percent 
change from officers five tickets below their peers, would be difficult for a 
driver to perceive. One would need to receive more than 250 tickets from an 
officer to reject the hypothesis that their ticket writing rate is not constant 
between the two periods.

The twenty-five percent increase is significant because it represents a systemic 
shift in activity, and at issue is the quality of that activity. A quota system 
is illegal because the micromanagement of police officers reduces discretion. A 
de facto quota system, based on a "ticket race", serves the same purpose. 
History has revealed that the indiscriminate enforcement of minor violations can 
provoke confrontations between police and the community and erode the public 
trust necessary for combating serious offenses, which is in direct opposition to 
the officer's mission. In extreme cases, a ticket can ruin entire families by 
triggering the parole violation, incarceration or deportation of an otherwise 
upstanding resident.

Yet performance goals need not micromanage officers. In the landmark opinion 
that rejected the implementation of NYPD's Stop and Frisk procedure, the judge 
found the use of performance goals under Q4E "created pressure to carry out 
stops, without any system for monitoring the constitutionality of those stops. 
However, the use of performance goals in relation to stops may be appropriate, 
once an effective system for ensuring constitutionality is in place." 
[@scheidlin2013, pg. 17] We believe data analysis will be an important part of 
any system, and we see this work as contributing towards a conversation of how 
such an analysis could be performed.

## VI. References