---
title: "R Notebook"
output: html_notebook
---

```{r setup}
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
library(tidyverse)
library(bayesplot)
library(shinystan)
library(here)
```


Note: the model is able to fit correctly when measurement_sigma_absolute and measurement_sigma_relative are treated as parameters, but it is not very useful for real data - either measurement_sigma_relative has to be tightly constrained by prior or it grows large enough to accomodate data without accomodating the underlying spline and regulation


```{r}
regulated_model <- stan_model(file = here('stan','regulated.stan'))
```

```{r}
data <- simulate_regulated_spline(3,31,5,seq(1,31,by = 3), 0.2, 0.2, regulator_measured = TRUE, integrate_ode45 = FALSE)
print(data$true[c("degradation","sensitivity", "w","b_raw")])
data.frame(time = 1:data$observed$num_time, regulator = data$true$regulator_expression, target1 = data$true$expression[1,], target2 = data$true$expression[2,], target3 = data$true$expression[3,]) %>%
  gather("profile", "value",-time) %>%
  ggplot(aes(x=time, y = value, color = profile)) + geom_line()
```

```{r}
data <- simulate_regulated_spline(5,31,seq(1,31,by = 3), num_df = 5, measurement_sigma_absolute = 0.1, measurement_sigma_relative = 0.2, regulator_measured = TRUE, integrate_ode45 = FALSE)

fit <- sampling(regulated_model, data = data$observed, control = list(adapt_delta = 0.95))
evaluation_summary(rstan::extract(fit, c("coeffs","initial_condition","sensitivity","degradation", "intercept","w","b","sigma_relative_param","sigma_absolute_param")), data$true, printParamsResults = TRUE)
 summary(fit)$summary[,c("n_eff","Rhat")] %>% as.data.frame() %>% summarise(min_n_eff = min(n_eff),max_Rhat = max(Rhat))

fitted_regulator_plot(fit, data$observed)

for(t in 1:data$observed$num_targets) { 
  fitted_target_plot(fit, data$observed, target = t) %>% print()

}


```
```{r}
np_style <- parcoord_style_np(div_alpha = 0.6)

mcmc_parcoord(as.array(fit, pars = c("coeffs","degradation","sensitivity","initial_condition","w","b_raw","intercept_raw")), np = nuts_params(fit), np_style = np_style)

mcmc_parcoord(as.array(fit, pars = c("coeffs","initial_condition","b_raw","intercept_raw")), np = nuts_params(fit), np_style = np_style)


target <- 3
expression_data <- as.array(fit, pars = c("predicted_expression"))
target_data <- expression_data[,,seq(1,(dim(tt)[3]), by = 3) - 1 + target]
mcmc_parcoord(target1_data, np = nuts_params(fit), np_style = np_style)


mcmc_parcoord(as.array(fit, pars = c("predicted_regulator_expression")), np = nuts_params(fit), np_style = np_style)

```

# TODO check alternatives to ensuring positivity

```{r}
points <- 1:100
df <- 7
basis <- bs(points, df = df, intercept = TRUE)
n_samples <- 10
rand_coeff <- array(rnorm(df*n_samples), c(df, n_samples))
raw_points <- basis %*% rand_coeff
min_add <- array(apply(raw_points, 2, min), c(n_samples, length(points))) %>% t
min_plot <- ggmatplot(points, raw_points - min_add, main_geom = geom_line(alpha = 0.5)) + ggtitle("Min = 0")
exp_plot <- ggmatplot(points, log1p(exp(raw_points)), main_geom = geom_line(alpha = 0.5)) + ggtitle("log(1 + exp)") 
plot_grid(min_plot, exp_plot, ncol = 1)

```
